---
layout: default
title: "Working with the Household Grid"
nav_order: 4
parent: MCS
format:
  gfm:
    keep-yaml: true
---

- [Download the R script for this page](../purl/mcs-household_grid.R)
- [Download the equivalent Stata script for this page](../do_files/mcs-household_grid.do)

# Introduction

In this section, we describe the basics of using the household grid. Specifically, we show how to use the household grid to:

1.  Identify particular family members

2.  Create family-member specific variables

3.  Determine the relationships between non-cohort members within a family.

We use the following packages:

```{r}
#| warning: false
# Load Packages
library(tidyverse) # For data manipulation
library(haven) # For importing .dta files
```

```{r}
#| include: false
# setwd(Sys.getenv("mcs_fld"))
```

# Finding Mother of Cohort Members

To show how to perform 1 & 2, we use the example of finding natural mothers' smoking status at the first sweep. We load just four variables from the Sweep 1 household grid: `MCSID` and `APNUM00`, which together uniquely identify an individual, and `AHPSEX00` and `AHCREL00`, which contain information on the individual's sex and their relationship to the household's cohort member(s). `AHCREL00 == 7` identifies natural parents and `AHPSEX00 == 2` identifies females. Combining the two identifies natural mothers. Below, we use `count()` to show the different (observed) values for the sex and relationship variables. We also use the `filter()` function (which retains observations where the conditions are `TRUE`) to create a dataset containing the identifiers (`MCSID` and `APNUM00`) of natural mothers only; we will merge this with the smoking information shortly. `add_count(MCSID) %>% filter(n == 1)` is included as an interim step to ensure there is just one natural mother per family.[^1]

[^1]: Loading the `.dta` files into `R` with `haven::read_dta()` retains the dataset metadata, including variable names and labels, mainly by storing variables as `labelled` class objects. See the [`labelled` package help files](https://cran.r-project.org/web/packages/labelled/vignettes/intro_labelled.html) for more information on working with this metadata - for instance, converting `labelled` variables to standard `R` factor variables or replacing negative values (generally reserved in MCS data to indicate missingness) with `R`'s native `NA` value.

```{r}
df_0y_hhgrid <- read_dta("0y/mcs1_hhgrid.dta") %>%
  select(MCSID, APNUM00, AHPSEX00, AHCREL00) # Retains the listed variables

df_0y_hhgrid %>%
  count(AHPSEX00) # Tabulates each sex; AHPSEX00 does not record the sex of cohort members

df_0y_hhgrid %>%
  count(AHCREL00) # Tabulates each relationship to a cohort member

df_0y_mothers <- df_0y_hhgrid %>%
  filter(
    AHCREL00 == 7, # Keep natural parents...
    AHPSEX00 == 2 # ...who are female.
  ) %>%
  add_count(MCSID) %>% # Creates new variable (n) containing # of records with given MCSID
  filter(n == 1) %>% # Keep where only one recorded natural mother per family
  select(MCSID, APNUM00) # Keep identifier variables
```

Note, where a cohort member is part of a family (`MCSID`) with two or more cohort members, the cohort member will have been a multiple birth (i.e., twin or triplet), so familial relationships should apply to all cohort members in the family, which is why there is just one relationship (`[A-G]HCREL00`) variable per household grid file. This will change as the cohort members age, moving into separate residences and starting their own families.

# Creating a Mother's Smoking Variable

Now we have a dataset containing the IDs of natural mothers, we can load the smoking information from the Sweep 1 parent interview file (`mcs1_parent_interview.dta`). The smoking variable we use is called `APSMUS0A` and contains information on the tobacco product (if any) a parent consumes. We classify a parent as a smoker if they use any tobacco product (`mutate(parent_smoker = case_when(...))`).

```{r}
df_0y_parent <- read_dta("0y/mcs1_parent_interview.dta") %>%
  select(MCSID, APNUM00, APSMUS0A) # Retains only the variables we need

df_0y_parent %>%
  count(APSMUS0A)

df_0y_smoking <- df_0y_parent %>%
  mutate(parent_smoker = case_when(APSMUS0A %in% 2:95 ~ 1, # If APSMUS0A is integer between 2 and 95, then 1
                            APSMUS0A == 1 ~ 0)) %>% # If APSMUS0A is 1, then 0
  select(MCSID, APNUM00, parent_smoker)
```

Now we can merge the two datasets together to ensure we only keep rows in `df_0y_smoking` that appear in `df_0y_mothers`. We use `left_join()` to do this, with `df_0y_mothers` as the dataset determining the outputted rows, so that we have one row per identified mother.[^2] The result is a dataset with one row per family with an identified mother. We rename the `parent_smoker` variable to `mother_smoker` to clarify that it refers to the mother's smoking status.

[^2]: `left_join()` takes as arguments two data frames and retains only the rows in the first data frame, regardless of whether there is a match with the second. See [*Combining Data Across Sweeps*](https://cls-data.github.io/docs/mcs-merging_across_sweeps.html) for more discussion of the `*_join()` functions.

Below we also pipe this dataset into the `tabyl()` function (from `janitor`) to tabulate the number and proportions of mothers who smoke and those who do not.

```{r}
#| warning: false
# install.packages("janitor") # Uncomment if you need to install
library(janitor)
df_0y_mothers %>%
  left_join(df_0y_smoking, by = c("MCSID", "APNUM00")) %>%
  select(MCSID, mother_smoker = parent_smoker) %>%
  tabyl(mother_smoker)
```

# Determining Relationships between Non-Cohort Members

The household grids include another set of relationship variables besides `[A-G]HCREL00`. These vary in name slightly between sweeps: `[A-D]HPREL[A-Z]0` in `mcs[1-4]_hhgrid.dta`, `EPREL0[A-Z]00` in `mcs5_hhgrid.dta`, and `[F-G]HPREL0[A-Z]` in `mcs[6-7]_hhgrid.dta`. These variables can be used to identify the relationships between non-cohort member family members. Specifically, they record the person in the row's (ego) relationship to the person denoted by the column (alt); the letter `[A-Z]` in the variable name corresponds to the alt's `[A-D]PNUM00`. For instance, the variable `AHPRELB0` denotes the relationship of the person in the row to the person in the same family with `APNUM00 == 2`. Below, we extract a small set of data from the Sweep 1 household grid to show this in action.

```{r}
df_0y_hhgrid_prel <- read_dta("0y/mcs1_hhgrid.dta") %>%
  select(MCSID, APNUM00, matches("AHPREL[A-Z]0"))

df_0y_hhgrid_prel %>%
  filter(MCSID == "M10001N") %>% # To look at just one family
  select(APNUM00, AHPRELA0, AHPRELB0, AHPRELC0) # To look at first few relationship variables
```

There are seven members in this family, one of whom is a cohort member (`APNUM00 == 100`). `APNUM00`'s 1 and 2 are the (natural) parents, and `APNUM00`'s 3-6 and 100 are the (natural) children. The relationship variables show that `APNUM00`'s 1 and 2 are married, and `APNUM00`'s 3-7 are siblings (`AHPRELC0 == 11 [Natural brother/sister]`) and biological offspring of `APNUM00`'s 1 and 2 (`AHPREL[A-B]0 == 3 [Natural son/daughter]`). Note the symmetry in the relationships. Where, `APNUM00 == 1`, `AHPRELC0 == 7 [Natural Parent]` and where `APNUM00 == 3`, `AHPRELA0 == 3 [Natural son/daughter]`.

If we want to find the particular person occupying a specific relationship for an individual (e.g., we want to know the `[A-G]PNUM00` of the person's partner), we need to reshape the data into long-format with one row per ego-alt relationship within a family. For instance, if we want to find each person's spouse (conditional on one being present), we can do the following:[^3]

[^3]: For more on reshaping data, see [*Reshaping Data from Long to Wide (or Wide to Long)*](https://cls-data.github.io/docs/mcs-reshape_long_wide.html) for more discussion of the `*_join()` functions.

```{r}
df_0y_hhgrid_prel %>%
  pivot_longer(cols = matches("AHPREL[A-Z]0"),
               names_to = "alt",
               values_to = "relationship") %>%
  mutate(APNUM00_alt = match(str_sub(alt, -2, -2), LETTERS)) %>% # Creates alt's PNUM00 by matching penultimate letter to position in alphabet
  filter(relationship == 1) %>% # Keep where husband or wife
  select(MCSID, APNUM00, partner_pnum = APNUM00_alt)
```

# Coda

This only scratches the surface of what can be achieved with the household grid. The `mcs[1-7]_hhgrid.dta` files also contain information on cohort-member and family-member's dates of birth, which can be used to, for example, identify the number of resident younger siblings, determine maternal and paternal age at birth, and so on.

# Footnotes
