---
layout: default
title: "Combining Data Within A Sweep"
nav_order: 6
parent: MCS
format: gfm
---

- [Download the R script for this page](../purl/mcs-merging_within_sweep.R)
- [Download the equivalent Stata script for this page](../do_files/mcs-merging_within_sweep.do)

# Introduction

In this section, we show how to merge, collapse and reshape the various data structures within a given sweep of the Millennium Cohort Study (MCS) to create a dataset at the cohort member-level (i.e., one row per cohort member). This is likely to be the most useful data structure for most analyses, but similar principles can be applied to create other structures as needed (e.g., family-level datasets).

We use the following packages:

```{r}
#| warning: false
# Load Packages
library(tidyverse) # For data manipulation
library(haven) # For importing .dta files
```

```{r}
#| include: false
# setwd(Sys.getenv("mcs_fld"))
```

# Data Cleaning

We create a small dataset that contains information from Sweep 2 on: family country of residence, cohort member's ethnicity, whether any parent reads to the child, the warmth of the relationship between the parent and the child, family social class (National Statistics Socio-economic Classification; NS-SEC), and mother's highest education level. Constructing and combining these variables involves restructing the data in various ways, and spans the most common data engineering tasks involved in bringing together information from a single sweep.

The datasets we use in this tutorial are:

```{r}
family <- read_dta("3y/mcs2_family_derived.dta") # One row per family
cm <- read_dta("3y/mcs2_cm_derived.dta") # One row per cohort member
parent <- read_dta("3y/mcs2_parent_derived.dta") # One row per parent (responding)
parent_cm <- read_dta("3y/mcs2_parent_cm_interview.dta") # One row per parent (responding) per cohort member
hhgrid <- read_dta("3y/mcs2_hhgrid.dta") # One row per household member
```

We begin with the simplest variables: cohort member's ethnicity and family country of residence. Cohort member's ethnicity is stored in a cohort-member level dataset already (`mcs2_cm_derived`), so it does not need further processing. Below we rename the relevant variables and select it along with the cohort member identifiers, `MCSID` and `BCNUM00`.

```{r}
df_ethnic_group <- cm %>%
  select(MCSID, BCNUM00, ethnic_group = BDC08E00) # Retains the listed variables, renaming BDC08E00 as ethnic_group

df_ethnic_group
```

Family country of residence is stored in a family-level dataset (`mcs2_family_derived`). This also does not need any further processing at this stage. Later when we merging this data with `df_ethnic_group`, we perform a 1-to-many merge, so the data will be automatically repeated for cases where there are multiple cohort members in a family.[^1]

[^1]: It is also possible to expand a family level dataset so that it has as many rows as there are cohort-members in the family. `mcs2_family_derived.dta` contains a variable, `BDNOCM00`, with this information that can be used with the `tidyverse` function `uncount(BDNOCM00)` to achieve this. (The dataset `mcs_longitudinal_family_file` contains a variable `NOCMHH` which holds similar information.)

```{r}
df_country <- family %>%
  select(MCSID, country = BACTRY00)
```

Next, we create a variable that indicates whether *any* parent reads to the cohort member; in other words, we create a summary variable using data from individual parents. The `mcs2_parent_cm_interview` dataset contains a variable for the parent's reading habit to a given child (`BPOFRE00`). We first create a binary variable that indicates whether the parent reads to the cohort member at least once a week, and then create a summary variable indicating whether any (interviewed) parent reads (`max(parent_reads)`) by collapsing the data using `summarise()` on a [grouped data frame](https://r4ds.hadley.nz/data-transform.html#groups) (`group_by(MCSID, BCNUM00)`) to ensure this is calculated per cohort member. The result is a dataset with one row per cohort member with data on whether any parent reads to them.[^2]

[^2]: Below, for simplicity, we drop any rows with missing values (`drop_na()` step). Proper analyses may opt to use a different rule, which may require merging in other information (e.g., setting the value to missing unless all resident parents have been interviewed and provided a valid response).

```{r}
df_reads <- parent_cm %>%
  select(MCSID, BPNUM00, BCNUM00, BPOFRE00) %>%
  mutate(parent_reads = case_when(between(BPOFRE00, 1, 3) ~ 1, # Create binary variable for reading habit
                                  between(BPOFRE00, 4, 6) ~ 0)) %>%
  drop_na() %>% # Drops rows with any missing value (ensures we get a value where at least 1 parent gave a valid response).
  group_by(MCSID, BCNUM00) %>% # Groups the data so summarise() is performed per cohort member.
  summarise(parent_reads = max(parent_reads), # Calculates maximum value per cohort member
            .groups = "drop") # Removes the grouping from the resulting dataframe.

df_reads
```

We next show a different way of using `mcs2_parent_cm_interview`, reshaping the data from long to wide so that we have one row per cohort member. As an example, we create separate variables for whether the responding parent has a warm relationship with the cohort member (`BPPIAW00`), one using responses from the main carer and one from the secondary carer. As `mcs2_parent_cm_interview` have one row per parent x cohort member combination, we first create a variable indicating which parent is which (using information from `BELIG00`), and then reshape the warmth variable from [long to wide using `pivot_wider()`](https://r4ds.hadley.nz/data-tidy.html#widening-data).

```{r}
df_warm <- parent_cm %>%
  select(MCSID, BCNUM00, BELIG00, BPPIAW00) %>%
  mutate(var_name = ifelse(BELIG00 == 1, "main_warm", "secondary_warm"),
         warmth = case_when(BPPIAW00 == 5 ~ 1,
                           between(BPPIAW00, 1, 6) ~ 0)) %>%
  select(MCSID, BCNUM00, var_name, warmth) %>%
  pivot_wider(names_from = var_name, # The new variables are named main_warm and secondary_warm...
              values_from = warmth) # ... and take values from "warmth"
```

Next, we show an example of creating family level data using data from individual parents; in this case, a variable for family social class (NS-SEC) using data from `mcs2_parent_derived`. As `mcs2_parent_derived` a parent level dataset, we take the minimum of parents' NS-SEC (`BDD05S00`) within a family (lower values of `BDD05S00` indicate higher social class).

```{r}
df_nssec <- parent %>%
  select(MCSID, BPNUM00, parent_nssec = BDD05S00) %>%
  mutate(parent_nssec = if_else(parent_nssec < 0, NA, parent_nssec)) %>% # Negative values denote various forms of missingness.
  drop_na() %>%
  group_by(MCSID) %>%
  summarise(family_nssec = min(parent_nssec))

```

Finally, we create a variable for the mother's highest education level using the `mcs2_parent_derived` dataset. This involves merging in relationship information from the household grid and subsetting the rows so we are left with data for mothers only (see [*Working with the Household Grid*](https://cls-data.github.io/docs/mcs-household_grid.html). We separately filter the household grid for mothers only (`BHCREL00 == 7` \[Natural Parent\] and `BHPSEX00 == 2` \[Female\]) and select the highest education level variable (`BDDNVQ00`) from the `mcs2_parent_derived` dataset. We then merge the datasets together use `right_join()`, which in this case, gives a row for every mother in the dataset, regardless of whether they have education data or not (`right_join()` fills variables with `NA` where [the retained row does not have a match](https://r4ds.hadley.nz/missing-values.html#sec-missing-implicit)).[^3]

[^3]: More detail on merging with `right_join()` (and other `*_join()` variants) is provided in [*Combining Data Across Sweeps*](https://cls-data.github.io/docs/mcs-merging_across_sweeps.html), as well as [Chapter 19 of the R for Data Science textbook](https://r4ds.hadley.nz/joins.html#sec-mutating-joins).

```{r}
df_mother <- hhgrid %>%
  select(MCSID, BPNUM00, BHCREL00, BHPSEX00) %>%
  filter(between(BPNUM00, 1, 99),
         BHCREL00 == 7,
         BHPSEX00 == 2) %>%
  distinct(MCSID, BPNUM00) %>%
  add_count(MCSID) %>%
  filter(n == 1) %>%
  select(MCSID, BPNUM00)

df_parent_edu <- parent %>%
  select(MCSID, BPNUM00, mother_nvq = BDDNVQ00)

df_mother_edu <- df_parent_edu %>%
  right_join(df_mother, by = c("MCSID", "BPNUM00")) %>%
  select(-BPNUM00)
```

# Merging the Datasets

Now we have cleaned each variable, we can merge them together. The cleaned datasets are either at the family level (`df_country`, `df_nssec`, `df_mother_edu`) or cohort member level (`df_ethnic_group`, `df_reads`, `df_warm`). We begin with `df_ethnic_group` as this has all the cohort members participating at Sweep 2 in it. We then use `left_join()` to merge in other data so original rows are kept (and no more are added). To merge with a family-level dataset, we use `left_join(..., by = "MCSID")` as `MCSID` is the unique identifier for each cohort member. For the cohort member level datasets, we use `left_join(..., by = c("MCSID", "BCNUM00"))` as the combination of `MCSID` and `BCNUM00` uniquely identifies cohort members.

```{r}
df_ethnic_group %>%
  left_join(df_country, by = "MCSID") %>%
  left_join(df_reads, by = c("MCSID", "BCNUM00")) %>%
  left_join(df_warm, by = c("MCSID", "BCNUM00")) %>%
  left_join(df_nssec, by = "MCSID") %>%
  left_join(df_mother_edu, by = "MCSID")
```

# Footnotes
